---
title: "StatisticalInference - Peer Graded Assignment"
author: "Pete Petersen III"
date: "6/17/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(ggplot2)
library(tinytex)

```

## Part I  

Part 1: Simulation Exercise

Overview

The task is to investigate the exponential distribution in R and compare it with the Central Limit Theorem. The exponential distribution will be simulated in R with rexp(n,lambda) where lambda is the rate parameter. The mean of exponential distribution and the standard deviation are both 1/lambda where lambda = 0.2, and distribution of averages of 40 exponentials and will perform 1000 simulations.


Simulations: Include English explanations of the simulations you ran, with the accompanying R code. Your explanations should make clear what the R code accomplishes.
```{r}

  lambda <- 0.2
  set.seed(11)
  n <- 40

  simulation <- replicate(1000, rexp(n, lambda))
  mean_simulation <- apply(simulation, 2, mean)


```

Sample Mean versus Theoretical Mean: Include figures with titles. In the figures, highlight the means you are comparing. Include text that explains the figures and what is shown on them, and provides appropriate numbers.

```{r}

  sampleMean <- mean(mean_simulation)
  simulationmean <- mean(simulation)
  theoretical_mean <- (1/lambda)

```



```{r density, echo=FALSE, fig.width=5,fig.height=5,fig.cap="Fig. 1 Simulation Distribution snd Density"}
# Histogram shows differences
  g<- ggplot(data.frame(mean_simulation),
             aes(x=mean_simulation))
  g<- g + ggtitle("1000 Simulation of size = 40")
  g<- g + geom_histogram(binwidth = 0.3,color = "black",
                         fill="beige",
                         aes(y = ..density..))
  g<- g + geom_density(colour="red",size=1) +
      labs(x="Means",y="Density")
  print(g)
```

Sample Variance versus Theoretical Variance: Include figures (output from R) with titles. Highlight the variances you are comparing. Include text that explains your understanding of the differences of the variances.

```{r}
# sample deviation & variance
sample_sd <- sd(mean_simulation)
sample_var <- sample_sd^2
# theoretical deviation & variance
theoretical_sd <- (1/lambda)/sqrt(n)
theoretical_var <- ((1/lambda)*(1/sqrt(n)))^2
```

Distribution: Via figures and text, explain how one can tell the distribution is approximately normal.

  
```{r figs, echo=FALSE, fig.width=5,fig.height=5,fig.cap="Fig. 2 Distribution Analysis"}
g <-ggplot(data.frame(mean_simulation),aes(x=mean_simulation))
g <- g + ggtitle("Simulation Distribution: \n Normal vs Sample Density")
g <- g + geom_histogram(binwidth = 0.3,colour="black",fill="beige",aes(y = ..density..))
g <- g + geom_density(colour="blue",size=1)+labs(x="Mean Values",y="Density")
g <- g + stat_function(fun = dnorm, 
                     args = list(mean = 1/lambda,
                             sd=1/lambda/sqrt(n)),
                     colour="red",size=1,linetype=1)
g <- g +geom_vline(xintercept = theoretical_mean, size=1, colour="red") + 
        geom_vline(xintercept = sampleMean, size=1, colour="blue")  

print(g)
```
The above plot indicated that density curve is similar to normal distribution curve.

Q-Q Normal Plot also indicates the normal distribution
```{r}
qqnorm(mean_simulation)
 qqline(mean_simulation, col = "magenta")
```